---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
In order to complete all the analysis required in this assignment the dplyr 
library is required.

```{r chunk_name, message=FALSE}
library(dplyr);
```

The first thing that needs to be done is to unzip the activity.zip file which
contains the data which will be used in this assignment.

```{r}
unzip("activity.zip", overwrite = TRUE);
```

The next step is to get an understanding of what the data in this file looks
like. To do this the file will be read and then the first 10 rows will be 
displayed.

```{r}
activityData <- read.csv("activity.csv");
head(activityData, 10);
```

## What is mean total number of steps taken per day?
To determine the mean total number of steps taken per day the mean will be taken
of the data set grouped by day. 

```{r}
activityData_byDayMean <- activityData %>% group_by(date) %>% summarise_each(funs(mean))
activityData_byDayMean
```

Now that the mean is calculated by day, a histogram can be created to look at 
the distribution of steps per day.

```{r}
hist(activityData_byDayMean$steps, main="Steps Histogram", ylab="Frequency", xlab="# of Steps per 5 mins", col="lightblue", labels=TRUE, ylim=c(0,20))
```

From the histogram above it is clear that the highest level of frequency is in the 30-50 steps per 5 min period. 

Next a summary of the data will be considered with the focus being specifically on the values in steps.

```{r}
summary(activityData_byDayMean)
```

With the NA values being excluded the mean for steps per day is `r format(round(mean(activityData_byDayMean$steps, na.rm=TRUE),4), nsmall=4)`. The median is `r format(round(median(activityData_byDayMean$steps, na.rm=TRUE),4), nsmall=4)`

## What is the average daily activity pattern?
To answer the question of what the average daily activity pattern is, the data must
first be grouped by interval.

```{r}
activityData_byIntMean <- activityData %>% group_by(interval) %>% summarise_each(funs(mean(., na.rm=TRUE)))
activityData_byIntMean
```

With the data now grouped, a time series plot can be created showing the average number of steps over time.
This provides a visual of the individuals activity levels throughout the day.

```{r}
plot.ts(x=activityData_byIntMean$interval, y=activityData_byIntMean$steps, ylim=c(0,210), type="l", main="Daily Activity Pattern", xlab="5 min Interval", ylab="Average # of Steps")
```

## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
